A lottery model of density-dependent selection

### 2017-10-02 manuscript.tex
latest

\title{A lottery model of density-dependent selection in evolutionary genetics}
%\author{Jason Bertram $^{1,\ast}$ \\ 
%Joanna Masel $^{1}$}

@@

Density-dependent selection was one of the first topics to explicitly connect ecology and evolution \citep{travis_2013}.

The stated aim of MacArthur's famous $r$/$K$ scheme was to extend the existing population-genetic treatment of selection to account for population density \citep{macarthur_1962}. Few attempts have been made to develop it further along these lines.

 For example, MacArthur's famous $r$/$K$ scheme \citep{macarthur_1962,macarthur_1967} is now almost exclusively known as a framework for understanding life-history traits, and judged on its failure in that role \citep{pianka_1970,stearns_1977,boyce_1984,reznick_2002}.

It is common practice in population genetics to assign constant selection coefficients to genotypes describing the change in their relative frequencies over time.

This simplified treatment of selection facilitates explicit time-dependent treatment of genotype frequencies, and can be justified over sufficiently short time intervals \citep[p. 276]{ewens_2004}. The emphasis of population genetics is to infer past selection, migration and demographic change given a sample of nucleotide sequences, or to predict how allele frequencies change over time based on their relative fitness effects together with population structure, genetic drift and linkage. The resulting picture of evolution excludes basic elements of the ecological underpinnings of selection, including density dependence, and how selection affects population size. This complicates the inference of past selection, because demographic changes can look genealogically very similar to selective frequency changes \citep{barton_1998}.

By contrast, models of phenotypic trait evolution use absolute fitness functions to describe how some traits of interest affect survival and reproduction in particular ecological scenarios \citep{metz_1992,diekmann_2004}. These fitness functions can be quite problem-specific and often only account for a few traits at a time. The emphasis here is on the conditions for invasion from low frequencies and co-existence, rather than frequency or abundance trajectories over time. For instance, adaptive dynamics uses ``invasion fitness'' to explore the consequences of eco-evolutionary feedbacks \citep{diekmann_2004}.

@@

The first feature is that the lottery representation of competition is  particularly concise. Mature individuals (``adults'') each require their own territory, whereas newborn individuals (``propagules'') disperse to, and subsequently compete for, territories made available by the death of adults. Territorial contest among propagules leaves a single victorious adult per territory, the victor chosen at random from the propagules present, with probabilities weighted by a coefficient for each type representing competitive ability, akin to a lottery \citep{sale_77}. By comparison, coefficients for the pairwise effects of types on each other (e.g. the $\alpha$ coefficients in the generalized Lotka-Volterra equations and the associated concept of ``$\alpha$-selection''; \citealt{gill_1974,case_1974,joshi_2001}), or  explicit resource consumption \citep{tilman_1982}, are much more complicated. The second feature is the close connection between the lottery model and one of the foundational models of population genetics, the Wright-Fisher model of genetic drift, which we discuss further below.

>>>>

\title{A lottery model of density-dependent selection}
%\author{Jason Bertram $^{1,\ast}$ \\ 
%Joanna Masel $^{1}$}

@@

One of the simplest and most widely used models of selection is to assign genotypes constant selection coefficients describing how their relative proportions change over time. This model is common in population genetics, where its simplicity is useful for analyzing the detailed time course of genotypic frequencies.

Yet selection does not only change relative genotypic proportions. Ultimately, it involves the births and deaths of individuals, and is therefore intimately connected to changes in absolute abundance \cite{evolution_demography_stuff}. This immediately poses a challenge: any description of evolution grounded in abundances needs to specify how population density is regulated, as well as how fitness depends on density, which may differ between genotypes (density-dependent selection). As a result, the early density-dependent selection literature was among the first to explicitly grapple with the interplay between ecology and evolution \citep{travis_2013}.

Much of the work on density-dependent selection grew out of MacArthur's $r$/$K$-selection scheme \citep{macarthur_1962,macarthur_1967} in which $r$-selection favors faster growth at low density, whereas $K$-selection favors the ability to keep growing at high density. Although $r$ and $K$ are named after the common ``carrying capacity'' parameterization of the logistic model, the concepts of $r$- and $K$-selection are not restricted to that model \citep{macarthur_1967}. Nevertheless, much of the density-dependent selection literature has adopted variants of the logistic model, often in the setting of Mendelian inheritance \cite{roughgarden etc}.

with genotype-specific $K$ values parameterizing growth ability at high density. A closely related approach is to add Lotka-Volterra-type inter-genotypic competition terms with $\alpha$ competition coefficients, which adds ``$\alpha$-selection'' to the mix \citealt{gill_1974,case_1974,joshi_2001}.

 variant that allows for inter-genotypic versus intra-genotypic

or  explicit resource consumption \citep{tilman_1982}, are much more complicated.

$r$/$K$ scheme \citep{macarthur_1962,macarthur_1967} is now almost exclusively known as a framework for understanding life-history traits, and judged on its failure in that role \citep{pianka_1970,stearns_1977,boyce_1984,reznick_2002}.

@@

The first feature is that the lottery representation of competition is  particularly concise. Mature individuals (``adults'') each require their own territory, whereas newborn individuals (``propagules'') disperse to, and subsequently compete for, territories made available by the death of adults. Territorial contest among propagules leaves a single victorious adult per territory, the victor chosen at random from the propagules present, with probabilities weighted by a coefficient for each type representing competitive ability, akin to a lottery \citep{sale_77}.

The second feature is the close connection between the lottery model and one of the foundational models of population genetics, the Wright-Fisher model of genetic drift, which we discuss further below.
